---
title: "DIET COMPOSITION ANALYSIS"
author: "Nitsa"
date: "2023-02-05"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE, message=FALSE}
# LOAD PACKAGES
library(tidyverse)
library(knitr)
library(vegan)
library(dplyr)
library(EcolUtils)
library(RVAideMemoire)
library(ecodist)
```


```{r, message=FALSE, include = FALSE, echo=FALSE}
# UPLOAD DATA
DIET <- read_csv("DIET_DATA.csv")

# SUM PREY GROUP FREQUENCY & DRY MASS PER INDIVUDAL
PREY_DIET <- DIET %>%
  group_by(Sample_ID, PREY_GROUP) %>%
  mutate(prey_n = n(),
         prey_dm = sum(Dry_mass)) %>%
  distinct_at(vars(Sample_ID, PREY_GROUP), .keep_all = TRUE) %>%
  select(-c(Dry_mass)) %>%
  mutate(SEASON_PREDATOR = paste(Occasion, Species, sep = "_")) %>% # predictor variable
  select(-c(prey_n, Occasion))  %>%
  mutate(Group = Sample_ID) %>%
  select(Group, SEASON_PREDATOR, year, season, Sample_location, Species, PREY_GROUP, prey_dm) %>%
  mutate(prey_dm_4 = prey_dm^(1/4)) %>%
  select(-c(prey_dm)) %>%
  as.data.frame() %>%
  select(-c(Sample_ID))

# COUNT NUMBER OF DIETS
SCA_SAMPLE_SIZE <- PREY_DIET %>%
  select(Group, Species, PREY_GROUP) %>%
  distinct_at(vars(Group), .keep_all = TRUE) %>%
  group_by(Species) %>%
  mutate(SCA_N = n_distinct(Group)) %>%
  mutate(EMPTY = case_when(
    PREY_GROUP == "EMPTY" ~ "1",
    PREY_GROUP != "EMPTY" ~ "0")) %>%
  mutate(EMPTY = as.numeric(EMPTY)) %>%
  group_by(Species) %>%
  mutate(EMPTY_N = sum(EMPTY)) %>%
  mutate(PROP_EMPTY = EMPTY_N/SCA_N ) %>%
  distinct_at(vars(Species), .keep_all = TRUE) %>%
  select(Species, SCA_N, EMPTY_N, PROP_EMPTY)

# REMOVE
rm(SCA_SAMPLE_SIZE)

# PIVOT WIDER
WIDE_DATA <- PREY_DIET %>%
  filter(PREY_GROUP != "EMPTY") %>%
  pivot_wider(names_from = PREY_GROUP, values_from = prey_dm_4, values_fill = 0) 

# PREDICTOR VARIABLES
DIET_PREDICT <- WIDE_DATA %>% 
  select(Group, year, season, Sample_location, Species, SEASON_PREDATOR)

# MATRIX
DIET_MATRIX <- WIDE_DATA %>% 
  select(-c(Species, year, season, Sample_location, SEASON_PREDATOR)) %>%
  column_to_rownames("Group") %>%
  as.matrix()

# REMOVE
rm(WIDE_DATA)
```

```{r, include=FALSE, message = FALSE}
# "A rule of thumb: stress < 0.05 provides an excellent representation in reduced dimensions, < 0.1 is great, <0.2 is good/ok, and stress < 0.3 provides a poor representation."
# metaMDS repeats algortihim about 20 different times -- trying to find the smallest stress value 
# stress = the amount of distortion that happens when you take highly dimensional data and smush it down into 2 dimensions

# BRAY-CURTIS DISTANCE MATRIX
set.seed(02091997)
DIET_DIST_DM <- vegdist(DIET_MATRIX, method="bray")

# NMDS
set.seed(02091997)
DIET_MDS_DM <- metaMDS(DIET_DIST_DM, distance = "bray", autotransform = FALSE) 

# STRESS LEVEL
DIET_NMDS_STRESS <- DIET_MDS_DM$stress

# DATAFRAME
DIET_DIST <- DIET_MDS_DM %>%
  scores() %>%
  as_tibble(rownames = "Group")

# MERGE DATA
META_DIET_MDS_DM <- inner_join(DIET_PREDICT, DIET_DIST)
write.csv(META_DIET_MDS_DM, "\\NMDS2.csv", row.names=FALSE)
```

# TEST FOR DIFFERENCES IN DIET COMPOSITION BETWEEN GROUPS
## PERMANOVA 
```{r, echo = FALSE, include = FALSE}
#### ADONIS2 ####
# DIFFERENCES BETWEEN PREDATOR GROUPS & SEASONS
PERMANOVA_DM <- adonis2(DIET_DIST_DM ~ 
                          META_DIET_MDS_DM$Species * META_DIET_MDS_DM$season + 
                          META_DIET_MDS_DM$season +
                          META_DIET_MDS_DM$year + 
                          META_DIET_MDS_DM$Sample_location,
                        permutations = 9999,
                        method = "bray")
PERMANOVA_DM

# NOTE: 9999 and 999 permutationss have similar results 
```

```{r, echpo = FALSE}
# PERMANOVA RESULTS
# W/ SAMPLE LOCATION
PERMANOVA_DM
```

# PAIRWISE COMPARISONS BETWEEN GROUPS
## PERMANOVA PAIRWISE
```{r, include = FALSE, echo = FALSE}
# ECOLE PACKAGE
library(ecole)
PAIR_PREDATOR <- permanova_pairwise(x = DIET_DIST_DM, 
                                    grp = DIET_PREDICT$SEASON_PREDATOR, 
                                    permutations = 9999,
                                    method = 'bray', 
                                    adj = 'bonferroni')

######## SEASONS -- WITHIN PREDATOR GORUP CONTRASTS ######## 
PAIR_PREDATOR_SEASON <- PAIR_PREDATOR %>%
  filter(##### BNT ##### 
         pairs == "MAY_2021_BNT vs AUG_2021_BNT" |
         pairs == "MAY_2021_BNT vs OCT_2021_BNT" |
         pairs == "AUG_2021_BNT vs OCT_2021_BNT" |
         pairs == "MAY_2022_BNT vs AUG_2022_BNT" |
         pairs == "MAY_2022_BNT vs OCT_2022_BNT" |
         pairs == "AUG_2022_BNT vs OCT_2022_BNT" |
         #####  MTS ##### 
         pairs == "MAY_2021_MTS vs AUG_2021_MTS" |
         pairs == "MAY_2021_MTS vs OCT_2021_MTS" |
         pairs == "AUG_2021_MTS vs OCT_2021_MTS" | 
         pairs == "MAY_2022_MTS vs AUG_2022_MTS" |
         pairs == "MAY_2022_MTS vs OCT_2022_MTS" |
         pairs == "AUG_2022_MTS vs OCT_2022_MTS")

######## PREDATOR GROUPS -- BETWEEN PREDATOR GORUP CONTRASTS ######## 
PAIR_PREDATOR_GROUPS <- PAIR_PREDATOR %>%
  filter(##### 2021 ##### 
         pairs == "MAY_2021_BNT vs MAY_2021_MTS" |
         pairs == "AUG_2021_BNT vs AUG_2021_MTS" |
         pairs == "OCT_2021_BNT vs OCT_2021_MTS" |
         #### 2022  ##### 
         pairs == "MAY_2022_BNT vs MAY_2022_MTS" |
         pairs == "AUG_2022_MTS vs AUG_2022_BNT" |
         pairs == "OCT_2022_MTS vs OCT_2022_BNT")

# REMOVE
rm(PAIR_PREDATOR_SEASON)
```

```{r, echo = FALSE}
# PAIRWISE COMPARISONS
PAIR_PREDATOR_GROUPS
```

# TEST FOR DIFFERENCES IN DISPERSION BETWEEN GROUPS
## BETADISPER
- Significant differences among groups (PERMANOVA or ANOSIM) may occur for either or both of two reasons (Warton et al. 2012): 1) Differences in the means (centroids) OR 2) Differences in the amount of dispersion of sample units around those centroids
- TEST TO -- conclude that, although composition differs among predator groups & occasions, it does so because of differences in location rather than dispersion.
```{r, include = FALSE, echo = FALSE}
# BETA DISPERSION -- SIGNFICANT DIFFERNCES IN DISPERSION BETWEEN GROUPS
perm.eg.betadisper <- betadisper(d = DIET_DIST_DM, group =  DIET_PREDICT$SEASON_PREDATOR,type = "centroid")
anova(perm.eg.betadisper)

# PARIWSIE DIFFERENCES IN DISPERSION
PRED_DISPER <- TukeyHSD(perm.eg.betadisper)
PRED_DISPER_DF <- PRED_DISPER$group %>% 
  as.data.frame()
  
# BOXPLOT
boxplot(perm.eg.betadisper)

######## SEASONS -- WITHIN PREDATOR GORUP CONTRASTS ######## 
PRED_DISPER_DF$Row_Name <- rownames(PRED_DISPER_DF)
DIST_PREDATOR_SEASON <- PRED_DISPER_DF %>%
  filter(##### BNT ##### 
         Row_Name == "MAY_2021_BNT-AUG_2021_BNT" |
         Row_Name == "OCT_2021_BNT-MAY_2021_BNT" |
         Row_Name == "OCT_2021_BNT-AUG_2021_BNT" |
         Row_Name == "MAY_2022_BNT-AUG_2022_BNT" |
         Row_Name == "OCT_2022_BNT-MAY_2022_BNT" |
         Row_Name == "OCT_2022_BNT-AUG_2022_BNT" |
         ##### MTS ##### 
         Row_Name == "MAY_2021_MTS-AUG_2021_MTS" |
         Row_Name == "OCT_2021_MTS-MAY_2021_MTS" |
         Row_Name == "OCT_2021_MTS-AUG_2021_MTS" | 
         Row_Name == "MAY_2022_MTS-AUG_2022_MTS" |
         Row_Name == "OCT_2022_MTS-MAY_2022_MTS" |
         Row_Name == "OCT_2022_MTS-AUG_2022_MTS") 

######## PREDATOR GROUPS -- BETWEEN PREDATOR GORUP CONTRASTS ######## 
DIST_PREDATOR_GROUPS <- PRED_DISPER_DF %>%
  filter(##### 2021 ##### 
         Row_Name == "MAY_2021_MTS-MAY_2021_BNT" |
         Row_Name == "AUG_2021_MTS-AUG_2021_BNT" |
         Row_Name == "OCT_2021_MTS-OCT_2021_BNT" |
         #### 2022  ##### 
         Row_Name == "MAY_2022_MTS-MAY_2022_BNT" |
         Row_Name == "AUG_2022_MTS-AUG_2022_BNT" |
         Row_Name == "OCT_2022_MTS-OCT_2022_BNT")

```

```{r, echo = FALSE}
# DISPERSION DIFFERENCES
DIST_PREDATOR_GROUPS
```

# NMDS PLOT
```{r, include = TRUE, message = FALSE, echo = FALSE}
# RE LEVEL OCCASION & PREDATOR GROUPS
# META_DIET_MDS_DM$Occasion <- ordered(META_DIET_MDS_DM$Occasion, levels = c("MAY_2021", "AUG_2021", "OCT_2021", "MAY_2022", "AUG_2022", "OCT_2022"))
META_DIET_MDS_DM$season <- ordered(META_DIET_MDS_DM$season, levels = c("May", "August", "October"))
META_DIET_MDS_DM$year <- ordered(META_DIET_MDS_DM$year, levels = c("2021", "2022"))

# META_DIET_MDS_DM <- META_DIET_MDS_DM %>% filter(Group != "T2080521251")?

# DEFINE MEAN CENTROID #
CENTROID_MEAN <- META_DIET_MDS_DM %>%
  group_by(year, season, Species) %>%
  summarize(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2))

CENTROID_MEDIAN <- META_DIET_MDS_DM %>%
  group_by(year, season, Species) %>%
  summarize(NMDS1 = median(NMDS1), NMDS2 = median(NMDS2))

##################### PLOTS ##################### 
###### SPECIES w/ 95% ELLIPSES + MEAN/MEDIAN CENTROIDS ###### 
META_DIET_MDS_DM %>%
    ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
    stat_ellipse(level = 0.95) +
    facet_grid(year~season)  +
    geom_point(data = CENTROID_MEAN, size = 3, shape = 21, color = "black", aes(fill = Species)) +
    geom_point(data = CENTROID_MEDIAN, size = 3, shape = 25, color = "black", aes(fill = Species)) +
    theme_minimal()

###### SPECIES W/ POINTS ###### 
META_DIET_MDS_DM %>%
    ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
    facet_grid(year~season)  +
    theme_minimal() +
    geom_point() 

###### SPECIES w/ 95% ELLIPSES + 40% ELLIPSES + MEDIAN CENTROID ###### 
NMDS_ELLIPSE <- META_DIET_MDS_DM %>%
    ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
    stat_ellipse(level = 0.95, linetype = 2) +
    stat_ellipse(level = 0.40) +
    facet_grid(year~season)  +
    theme_minimal() +
    geom_point(data = CENTROID_MEDIAN, size = 3, shape = 21, color = "black", aes(fill = Species)) + 
  labs(caption = str_wrap("Non-metric multidimensional scaling plots (NMDS) of brown trout and mottled sculpin diets. Dashed and solid ellipses show 95% and 40% confidence intervals, respectively. Points represent median centroids.")) +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        plot.caption = element_text(hjust = 0))

NMDS_POINTS <- META_DIET_MDS_DM %>%
    ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
    geom_point(aes(shape=Species, color=Species), size=1) +
    scale_shape_manual(values=c(1, 2))+
    facet_grid(year~season)  +
    theme_minimal() +
    ggtitle("") +
  stat_ellipse(level = 0.99, geom="polygon", alpha = 0.1, aes(fill = Species)) + 
  labs(caption = str_wrap("Non-metric multidimensional scaling plots (NMDS) of brown trout and mottled sculpin diets. Ellipses show 95% confidence interval. For improved visualization, eight outlier points have been excluded.")) +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        plot.caption = element_text(hjust = 0)) +
  coord_cartesian(xlim = c(-0.025, 0.025), ylim = c(-0.015, 0.015))

NMDS_SEASONS <- META_DIET_MDS_DM %>%
    ggplot(aes(x = NMDS1, y = NMDS2, color = season)) +
    geom_point(aes(shape=season, color=season), size=1) +
    facet_grid(year~Species)  +
    theme_minimal() +
    ggtitle("") +
  stat_ellipse(level = 0.99, geom="polygon", alpha = 0.1, aes(fill = season)) + 
  labs(caption = str_wrap("Non-metric multidimensional scaling plots (NMDS) of brown trout and mottled sculpin diets. Ellipses show 95% confidence interval. For improved visualization, eight outlier points have been excluded.")) +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        plot.caption = element_text(hjust = 0)) +
  coord_cartesian(xlim = c(-0.025, 0.025), ylim = c(-0.015, 0.015))


rm(CENTROID_MEAN, CENTROID_MEDIAN, NMDS_SEASONS)
```

```{r, fig.width=16, fig.height=9}
META_DIET_MDS_DM2 <- META_DIET_MDS_DM %>%
  mutate(SEASON = case_when(
    season == "May" ~ "Spring",
    season == "August" ~ "Summer", 
    season == "October"  ~ "Autumn")) %>%
  mutate(SEASON = factor(SEASON)) %>%
  mutate(SEASON = fct_relevel(SEASON,c("Spring","Summer", "Autumn"))) %>%
  mutate(YEAR = year) %>%
  mutate(YEAR = factor(YEAR)) %>%
  mutate(YEAR = fct_relevel(YEAR, c("2021","2022"))) %>%
  mutate(SPECIES = case_when(
    Species == "BNT" ~ "Brown trout",
    Species == "MTS" ~ "Mottled sculpin"))

META_DIET_MDS_DM2 %>%
    ggplot(aes(x = NMDS1, y = NMDS2, color = SPECIES)) +
    geom_point(aes(shape=SPECIES, color=SPECIES), size=3) +
    scale_shape_manual(values=c(1, 2))+
    facet_grid(YEAR~SEASON)  +
    theme_minimal() +
    ggtitle("") +
  stat_ellipse(level = 0.95, geom="polygon", alpha = 0.1, aes(fill = SPECIES)) + 
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        plot.caption = element_text(hjust = 0)) +
  coord_cartesian(xlim = c(-0.025, 0.025), ylim = c(-0.015, 0.015)) +
  theme(
        legend.title = element_blank(),
        legend.position = "bottom",
        legend.spacing.x = unit(0.8, 'cm'),
        legend.text = element_text(size=16,  
                                   color="black"),
        strip.text = element_text(size=18,
                                  color = 'black',
                                  face = "bold"),
         axis.title.x = element_text(size = 18,
                                   color = "black",
                                   face = "bold"),
        axis.text.x = element_text(size = 16,
                                   color = "black"),
        axis.title.y = element_text(size = 18,
                                   color = "black",
                                   face = "bold"),
        axis.text.y = element_text(size = 16,
                                   color = "black"),
        panel.grid.major = element_line(size = 0.2),
        panel.grid.minor = element_line(size = 0.2),
        panel.grid.major.x = element_blank()) + 
    guides(fill = guide_legend(nrow = 1)) +
  scale_color_manual(breaks = c("Brown trout", "Mottled sculpin"), 
                     labels = c("Brown trout", "Mottled sculpin"),
                     values=c("#ea5e00", "#bdbf01")) + 
    scale_fill_manual(breaks = c("Brown trout", "Mottled sculpin"), 
                     labels = c("Brown trout", "Mottled sculpin"),
                     values=c("#ea5e00", "#bdbf01")) 

# SAVE PLOT
ggsave("NMDS.jpeg", width = 16, height = 9, units = "in")
```



```{r}
NMDS_ELLIPSE
NMDS_POINTS
```

# PREY CONTRIBUTING TO DIFFERENCES
```{r, include = FALSE, echo = FALSE}
# SIMPER -- cumsum = proportion of dissimilarity between groups explained by that prey group (average/meandist)
(sim_dm <- with(DIET_PREDICT, simper(DIET_MATRIX, SEASON_PREDATOR, permutations = 9999)))
summary(sim_dm)

######## SEASONS -- WITHIN PREDATOR GORUP CONTRASTS ######## 
## BNT
MAY_AUG_BNT_2021 <- sim_dm$MAY_2021_BNT_AUG_2021_BNT %>%
  as.data.frame()
MAY_OCT_BNT_2021 <- sim_dm$MAY_2021_BNT_OCT_2021_BNT %>%
  as.data.frame()
AUG_OCT_BNT_2021 <- sim_dm$AUG_2021_BNT_OCT_2021_BNT %>%
  as.data.frame()
## MTS
MAY_AUG_MTS_2021 <- sim_dm$MAY_2021_MTS_AUG_2021_MTS %>%
  as.data.frame()
MAY_OCT_MTS_2021 <- sim_dm$MAY_2021_MTS_OCT_2021_MTS %>%
  as.data.frame()
AUG_OCT_MTS_2021 <- sim_dm$AUG_2021_MTS_OCT_2021_MTS %>%
  as.data.frame()

######## PREDATOR GROUPS -- BETWEEN PREDATOR GORUP CONTRASTS ######## 
## 2021
MAY_2021_BNTvMTS <- sim_dm$MAY_2021_BNT_MAY_2021_MTS %>%
  as.data.frame()
AUG_2021_BNTvMTS <- sim_dm$AUG_2021_BNT_AUG_2021_MTS %>%
  as.data.frame()
OCT_2021_BNTvMTS <- sim_dm$OCT_2021_BNT_OCT_2021_MTS %>%
  as.data.frame()

## 2022
MAY_2022_BNTvMTS <- sim_dm$MAY_2022_BNT_MAY_2022_MTS %>%
  as.data.frame()
AUG_2022_BNTvMTS <- sim_dm$AUG_2022_MTS_AUG_2022_BNT %>%
  as.data.frame()
OCT_2022_BNTvMTS <- sim_dm$OCT_2022_MTS_OCT_2022_BNT %>%
  as.data.frame()
```

```{r, message = FALSE, echo = FALSE}
# PREY CONTRIBUTIONS -- 2021
MAY_2021_BNTvMTS
AUG_2021_BNTvMTS
OCT_2021_BNTvMTS
```

```{r, message = FALSE, echo = FALSE}
# PREY CONTRIBUTIONS -- 2022
MAY_2022_BNTvMTS
AUG_2022_BNTvMTS
OCT_2022_BNTvMTS
```

# DIET COMP PLOTS
```{r, include=FALSE, echo = FALSE}
############# INDIVUDAL AVERAGE PROPORTION #############
# FORMAT DATA
IND_COMP <- DIET %>%
  filter(PREY_GROUP != "EMPTY") %>%
  select(year, season, Sample_ID, Species, PREY_GROUP, Dry_mass) %>%
  group_by(Sample_ID, PREY_GROUP) %>%
  mutate(sum_prey_group = sum(Dry_mass)) %>%
  distinct_at(vars(year, season, Sample_ID, PREY_GROUP), .keep_all = TRUE) %>%
  group_by(Sample_ID) %>%
  mutate(sum_total_prey = sum(sum_prey_group)) %>%
  mutate(prop_prey = sum_prey_group/sum_total_prey) %>%
  group_by(year, season, Species) %>%
  mutate(distinct_ID = n_distinct(Sample_ID)) %>%
  group_by(year, season, Species, PREY_GROUP) %>%
  mutate(sum_prop = sum(prop_prey)) %>%
  mutate(avg_prop = sum_prop/distinct_ID) %>%
  distinct_at(vars(year, season, Species, PREY_GROUP), .keep_all = TRUE) %>%
  select(year, season, Species, PREY_GROUP, avg_prop) %>%
  mutate(round_avg_prop = round(avg_prop, digits = 3))

# SPECIES MEANS
IND_COMP_MEAN <- IND_COMP %>%
  group_by(season, Species, PREY_GROUP) %>%
  summarise(mean_prey = mean(round_avg_prop))

# RELEVEL
IND_COMP$season <- ordered(IND_COMP$season, levels = c("May", "August", "October"))
IND_COMP$year <- ordered(IND_COMP$year, levels = c("2021", "2022"))

# PLOT
IND_COMP_PLOT <- ggplot(IND_COMP, aes(x=Species, y=avg_prop, fill=PREY_GROUP)) +
  geom_bar(stat="identity", aes(fill = PREY_GROUP)) +
  facet_grid(year ~ season) +
  theme_minimal() +
  labs(fill='Prey Group:') +
  labs(title = "Diet Composition of Brown Trout & Mottled Sculpin (Dry Mass of Prey)",
              subtitle = "average proportion of a given prey group in an individual's diet") + 
  ylab("Diet Composition") +
  theme(legend.position="bottom") 
```

```{r}
# DIET COMP PLOT
IND_COMP_PLOT

# WRITE CSV
write.csv(IND_COMP, "\\DIET_COMP.csv", row.names=FALSE)

```


